---
title: "Weight Lifting Prediction Analysis"
author: "Sérgio Henrique Barra"
date: "September 14, 2017"
output: html_document
---
## 1) Summary

This assignment is a course project from Practical Machine Learning course from
Coursera and Johns Hopkins University. The dataset is a Weight Lifting Exercise
(more info [here](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har)),
from Velloso, E., Bulling, A., Gellersen, H., Ugulino, W. and Fuks, H.

The dataset consists in data of 1.25kg-weight lifting of six male participants aged
between 20 and 28 years, evaluating how well a participant performed an 
Unilateral Dumbbell Biceps Curl. The exercise was repeated 10 times, in five different
styles of execution:  

* Exactly according to the specification (Class A),  
* Throwing the elbows to the front (Class B),  
* Lifting the dumbbell only halfway (Class C),  
* Lowering the dumbbell only halfway (Class D),  
* Throwing the hips to the front (Class E).

## 2) Loading data
```{r, echo = FALSE}
knitr::opts_chunk$set(message = FALSE)
```

Loading the necessary packages for the analysis:
```{r, message = FALSE}
library(caret)
library(parallel) # For parallel processing
library(doParallel)
library(randomForest)
library(pander)
```

Downloading, if needed, and reading the training and testing datasets.
```{r}
if (!(file.exists("pml-training.csv") & file.exists("pml-testing.csv"))) {
trainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

download.file(trainUrl, "./pml-training.csv")
download.file(testUrl, "./pml-testing.csv")
}

raw_training <- read.csv("./pml-training.csv", stringsAsFactors = FALSE)
testing <- read.csv("./pml-testing.csv", stringsAsFactors = FALSE)
```

## 3) Cleaning Data

The raw training data has **`r dim(raw_training)[1]`** observations of **`r dim(raw_training)[2]`** variables.

However, some variables have almost all NA or NULL values. Let's check which ones are those:
```{r}
NAsum <- sapply(raw_training, function(x) sum(is.na(x)))
NULLsum <- sapply(raw_training, function(x) sum(x == "", na.rm = TRUE))

tbNA <- table(NAsum)
tbNULL <- table(NULLsum)

df <- data.frame(tbNA, tbNULL)
names(df) <- c("NA in a column", "Frequency of NA", "NULL in a column", "Frequency of NA")
pander(df)
```

As we can see, there are **`r tbNA[2]`** columns with **`r names(tbNA)[2]`** missing values.
This represents **`r round(100*max(NAsum)/dim(raw_training)[1], 1)`%** of NA values in a column,
therefore being low valuable for building a machine learning model. Similarly,
this also happens with the NULL values. We can now remove these columns from the raw training dataset.
```{r}
NA_NULLind <- which(NAsum > 0|NULLsum > 0)

filtered_training <- raw_training[, -NA_NULLind]
```

For the models, we will consider that the first 7 variables (`r names(filtered_training[,1:7])`)
are not relevant. $X$ is unique and the other 6 variables are only for reference purposes.
```{r}
cleaned_training <- filtered_training[, 8:60]

print(names(cleaned_training))
```

Then, we need to format the integer columns into numeric and the $classe$ column
into factor.
```{r}
INTind <- sapply(cleaned_training, function(x) class(x) == "integer")
cleaned_training[, INTind] <- sapply(cleaned_training[, INTind], as.numeric)
cleaned_training[, 53] <- as.factor(cleaned_training[, 53])
```

## 4) Machine Learning Model

### 4.1)  Data splitting

We will split cleaned_training dataset into two: **training** and **validation**, where the latter
will represent **25%** of the dataset.
```{r}
set.seed(1)
inTrain <- createDataPartition(y = cleaned_training$classe, p = 0.75, list = F)
training <- cleaned_training[inTrain, ]
validation <- cleaned_training[-inTrain, ]
```

### 4.2) Random Forest Model

Now, we train the Random Forest model using parallel processing, with a 5 k-fold
cross-validation.
```{r rf_method, cache = TRUE}
cluster <- makeCluster(detectCores() - 1)
registerDoParallel(cluster)

start.time <- Sys.time()

set.seed(1)
ctrl <- trainControl(method = "cv", number = 5, allowParallel = TRUE)
mod_rf <- train(classe ~ ., data = training, trControl = ctrl, method = "rf")

elapsed_rf <- Sys.time() - start.time

stopCluster(cluster)
registerDoSEQ() # R returns to single-threaded processing

print(elapsed_rf)
```

## 4.3) Gradient Boosted Regression Model

We could also use GBM to build the model, under the same conditions of
parallel processing and resampling method:
```{r gbm_method, cache = TRUE}
cluster <- makeCluster(detectCores() - 1)
registerDoParallel(cluster)

start.time <- Sys.time()
set.seed(1)
ctrl <- trainControl(method = "cv", number = 5, allowParallel = TRUE)
mod_gbm <- train(classe ~ ., data = training, trControl = ctrl, method = "gbm", verbose = FALSE)
elapsed_gbm <- Sys.time() - start.time

stopCluster(cluster)
registerDoSEQ() # R returns to single-threaded processing

print(elapsed_gbm)
```

## 5) Validation

### 5.1) Random Forest
```{r}
pred_rf <- predict(mod_rf, validation)

confusionMatrix(pred_rf, validation$classe)
```

### 5.2) Generalized Boosted Regression
```{r, message = FALSE}
pred_gbm <- predict(mod_gbm, validation)

confusionMatrix(pred_gbm, validation$classe)
```

## 6) Model evaluation

Considering both models, we can see that although the GBM method ran faster than
Random Forest, it achieved a lower accuracy. Therefore, **we will choose the Random Forest
model for applying over the testing dataset**.

We can see below the importance of the first 15 variables considered by RF model:
```{r}
plot(varImp(mod_rf), top = 15, main = "Random Forest variable importance")
```

We can see that **roll_belt**, **pitch_forearm** and **yaw_belt** are the three
most important variables in the model.

## 7) Predicting testing classe variable

Now, using the built model to predict $classe$ in testing dataset:
```{r}
pred <- predict(mod_rf, testing)
print(pred)
```
